**Density Dependence**

### 2023-03-08 Rmd/report mac.Rmd
Minor updates; added some text.

`r format(Sys.time(), '%d/%m/%Y')`

Version: 1.0 (draft)

@@

**Stochastic projections at all** $F_{MSY}$'s from all scenarios

@@

@@

@@

Since the early 1980s, natural mortality ($M$) for mackerel has been fixed within the assessment model at 0.15, for all ages and all assessment years. This value was calculated based on estimates of total mortality derived from tagging data combined with catch data [@hamre1980biology]. M must, however, vary both by age and over time due to changes in abundance of predators, feeding conditions, disease etc. [@siegfried2009review]. Particularly any changes in mackerel distribution and growth will affect natural mortality. The impact of including age-varying natural mortality is expected to ....

>>>>

Version: 1.0 (07 March 2023)

Version: 1.1 (`r format(Sys.time(), '%d %B %Y')`)

@@

**Stochastic projections at all $F_{MSY}$'s from all scenarios**

Misspecification of Fmsy, i.e. using an Fmsy from one scenario and applying it to a different scenario has only limited impacts on the overall outcomes in terms of recruitment, biomass or yield. In the plot below, the scenarios for density dependence are shown in the columns and the Fmsy's are shown as the colours. For example a valid combination is the Base case with the red curves (Fmsy=0.33) or the M scenario with the green curve. Overall the misspecification does not results in noticeable differences.

@@

As a chech, we assessed the outcomes of the equilibrium curves from simulating over a wide F range, with the equilibrium values (points) in 2050 from the forward projections at different values of Fmsy (colours). The points fall on the equilibrium curve, thereby confirming the consistency between the two approaches.

@@

Estimating density-dependence from observations was only possible for mass-at-age and maturity age. Therefore, density dependence in natural mortality can only be explored from theoretical considerations, because current assessment does not include age or time varying natural mortality. M is likely to vary by age and time in relation to processes such as density dependence. While the recent expansion in distribution area and change in weight-at-age could be associated with changes in natural mortality. Therefore in the OMs we modelled M as a function of mass-at-age.

@@

Since the early 1980s, natural mortality ($M$) for mackerel has been fixed within the assessment model at 0.15, for all ages and all assessment years. This value was calculated based on estimates of total mortality derived from tagging data combined with catch data [@hamre1980biology]. M must, however, vary both by age and over time due to changes in abundance of predators, feeding conditions, disease etc. [@siegfried2009review]. Particularly any changes in mackerel distribution and growth will affect natural mortality.

Given the uncertainty in the assessment and the estimation of density dependence, it is likely that the Pretty Good Yield approach would ensure sustainable exploitation. Also, because we did not yhet consider stochastic variability in future developments in mass, maturity and natural mortality which are likely to obscure the impacts of density dependence. Variations in these processes due to changes in climate and range expansion are also likely to occur.

It is suggested that emprirical harvest control rules that respond to changes in productivity could be evaluated in contrast to the highly complex harvest rules based on SAM stock assessment models.
